A Systematic Approach to Model Predictive Controller Constraint Handling: Rigorous Geometric Methods

### 2011-04-28 article.tex
- Carl comments fix - added used packages to repo [just to be safe]

A rigorous and systematic approach to constraint management is proposed, building on the work of \citet{vinsonartoi} in interpreting constraint interactions. 
The method focuses on linear steady-state system models, and provides routines to obtain the following information; effects of constraint changes on the corresponding input or output constraints, feasibility checks for constraints, constraint-type information, specification of constraint-set size and optimal fitting of constraints within the desirable input or output space.

@@

%% ==== BODY ====
\section{Introduction}\label{sec:intro}
The ability of some MPC algorithms to impose constraints on the outputs or inputs of a system is their biggest selling point.
Just as the process model predicts the effect of the inputs on the outputs, the constraints imposed on the system are also interdependent (via the process model).

@@

Mathematical rigour and unambiguous language when specifying constraints were key design criteria.
Linear steady-state models are considered along with linear constraints on the inputs and outputs.
With the scope defined thus, the project can now focus exclusively on convex sets in the input and output space.

@@

\subsubsection{Constraint Formulation}
Constraints on the inputs ($\vect{u}$) or outputs ($\vect{y}$) of a system can be represented as sets of linear inequalities.
For the most general case, they can be expressed as follows;
\begin{align}
  \label{eq:gencons}
  A_u\vect{u}_k &\leq b_u \\ \notag
  A_y\vect{y}_k &\leq b_y \\ \notag
\end{align}
where $A_u$ and $A_y$ are coefficient matrices, and $b_u$ and $b_y$  the half-space offsets.

@@

\subsubsection{Process models}
The steady-state model of a linear system is the gain-matrix.
This is simply a matrix of constants in $\mathbb{R}$.
Models like these allow for matrix algebra when transforming between the input and output spaces (and vice versa).
Obtaining the AOS from the AIS is a linear transformation \citep[48-49,~166]{leung} corresponding to a matrix multiplication of the vertices of the AIS and the process model matrix.

@@

@@

\subsubsection{Constraints}\label{sec:mpccons}
The only validation of these constraints are against an outermost constraint set, often referred to as 'Engineering Limits'.
This set typically consists of physical limits on inputs and outputs, as well as limits that avoid damage to equipment.

Most commercial MPC packages allow only high and low limits to be imposed on outputs or inputs.

@@

\subsubsection{Overview}
%1966
The relative gain array (RGA) is one of the most used operability measures \citep[576]{luyben} and relates input and output interactivity.
Each element in the RGA ($\beta_{ij}$) is defined as the ratio of the steady-state gain ratios between input, $i$, and output, $j$, when all other inputs and outputs are constant.
Considering only the magnitudes of RGA elements can be misleading \citep[87]{skogestad}, as seemingly favourable pairings can be unfavourable due to the phase.
Therefore, both the magnitude and the phase need to be considered or, alternatively, norms such as the RGA-number can be used.
%1971
The Niederlinski index is another measure that only uses steady-state gains of the process model \citep[572-573]{luyben}.
Positive values of the Niederlinski index can correspond to unstable pairings and are therefore inconclusive.
For this reason, the index is considered necessary but not sufficient for stability \citep[445]{skogestad}.
Both the RGA and the Niederlinski index have the advantage of requiring little process information \citep[692]{marlin}.
%1985
\citet{artrdg} defined the RDG, which expanded on the RGA to include the effect of disturbances.
This expansion allowed for the calculation of MV saturation limits and system responses for specified load disturbances \citep{vinsonphd}. 
%1985
Singular value decomposition stems from the use of eigenvalues as an operability measure. 
The use of singular values are preferred as eigenvalues are a poor
measure of gain and can often be misleading \citep[75]{skogestad}.
Maximum and minimum singular values are mostly used to select controlled variables but can also be used as a performance measure along with the condition number \citepp{596}{luyben}{80-82}{skogestad}.

@@

The following spaces are of interest:
\begin{description}
  \item [Available Input Space (AIS);] the set of values that process      inputs can reach. 
The limits on these values are based on both physical limits (e.g.     valve openings) or process design values (e.g. flow-rates or temperatures). 
  \item [Achievable Output Space (AOS);] this is the set of values which the process outputs can obtain, given the AIS. 
The AIS maps to the AOS by means of a process model, $G$. 
Therefore, a point $u$ in the AIS corresponds to a point $y$ in the AOS via $y=G(u)$. 
  \item [Desired Output Space (DOS);] this represents the desired output values of the process. 
These values are typically based on operational and financial   parameters. 
  \item [Expected Disturbance Space (EDS);] all the values of the expected disturbances to the system. 
These values are translated back to the input space by means of a disturbance model.
  \item [Desired Input Space (DIS);] in the same manner that the AIS maps to the AOS, the DIS represents combined reverse mappings of the DOS and EDS.
The DIS is calculated as the inputs, $u$, that satisfy the model, i.e. $y=G(u,d)$ (with $y$ and $d$ from the DOS and EDS respectively).
\end{description}

@@

\subsection{Constraint types}\label{sec:contypes}
When constraints are considered, and specifically changes to them, the types of constraints present are important.
This is due to the direction of change allowed by each constraint type.
The following set of constraint types (based on the sources of constraints) are proposed to disambiguate the specification of constraints.
\begin{description}
  \item [Physical constraints] represent the physical limits of the system.
  Examples are tank levels, valve saturation limits and maximum heat duties.
  Allowed changes to these constraints are usually one-sided or to the inside of a range.
  \item [Quality constraints] are concerned with product quality, these constraints are typically determined by external requirements or standards.
  Strictly speaking, changes to these constraints are allowed in any direction.
  \item [Operational constraints] are the innermost constraint set used by the MPC controller.
  These constraints are typically changed using the operator interfaces of MPCs and are contained within their respective variables' physical and quality constraints.
  \item [Optimisation constraints] are those constraints that are not critical for the control of the plant, but rather the performance and efficiency of the plant as determined by external cost functions.
  Typically these external cost functions will represent a type of saving, e.g. energy savings or product give-away reduction.
\end{description}

@@

\section{Conclusions}\label{sec:conclusions}
From the proposed method of systematic constraint handling and the illustrative example, the conclusions can be drawn:
\begin{itemize}
  \item Checking constraints via the process model can identify constraints which are not attainable and should be revised.
  \item The constraints describing the AOS and DOS intersection represent the largest operating region for the given input and output constraints.
Reformatting these (typically linear combination) constraints by adding unmeasured variables to the process model allow for direct implementation in the high and low limit structure of commercial MPC interfaces.
  \item When constraint changes are made (during operation or design), checking changes to constraints via the process model allows for a measure of clamping (or relaxing) to be determined.
This serves to inform about the resulting effect of changing constraints and can guide design or operation decisions.
\end{itemize}

>>>>

A rigorous and systematic approach to constraint management is proposed, building on the work of \citet{vinsonartoi} in interpreting constraint interactions. 
The method focuses on linear steady-state system models, and provides routines to obtain the following information: effects of constraint changes on the corresponding input or output constraints, feasibility checks for constraints, constraint-type information, specification of constraint-set size and optimal fitting of constraints within the desirable input or output space.

@@

%% ==== BODY ====
\section{Introduction}\label{sec:intro}
Model predictive controllers use process models to predict the effect of the inputs of a system on its outputs.
The ability of some MPC algorithms to impose constraints on the outputs or inputs is their biggest selling point.
These constraints are also interdependent via the process model.

@@

Mathematical rigour and unambiguous language when specifying constraints were key design criteria.
Linear steady-state models are considered along with linear constraints on the inputs and outputs.
This paper will therefore focus only on convex sets in the input and output space.

@@

\subsubsection{Constraint Formulation}
Constraints on the inputs ($\vect{u}$) or outputs ($\vect{y}$) of a system can be represented as sets of inequalities.
For the most general linear case, they can be expressed as follows;
\begin{align}
  \label{eq:gencons}
  A_u\vect{u}_k &\leq b_u \\ \notag
  A_y\vect{y}_k &\leq b_y \\ \notag
\end{align}
where $A_u$ and $A_y$ are coefficient matrices, and $b_u$ and $b_y$  the half-space offsets.

@@

\subsubsection{Process models}
The steady-state model of a linear system is the gain-matrix.
This is simply a matrix of constants in $\mathbb{R}$.
With this model matrix, obtaining the output space from the input space is a linear transformation \citep[48-49,~166]{leung} corresponding to a matrix multiplication of the vertices of the input space and the process model matrix.
This same property holds when transforming back to the input space from the output space.

@@

Since the process model (gain-matrix) operates on and outputs deviation variables, a steady-state offset is added to produce absolute values.
This offset is calculated from the center of the input space and corresponds to the nominal point at which the process model is linearised.

@@

\subsubsection{Constraints}\label{sec:mpccons}
Most commercial MPC packages allow only high and low limits to be imposed on outputs or inputs.
The only validation of these constraints are against an outermost constraint set, often referred to as 'Engineering Limits'.
This set typically consists of physical limits on inputs and outputs, as well as limits that avoid damage to equipment.

@@

\subsubsection{Overview}
%1966
The relative gain array (RGA) is one of the most used operability measures \citep[576]{luyben} and relates input and output interactivity.
Each element in the RGA ($\beta_{ij}$) is defined as the ratio of the steady-state gain ratios between input, $i$, and output, $j$, when all other inputs and outputs are constant.
Considering only the magnitudes of RGA elements can be misleading \citep[87]{skogestad}, as seemingly favourable pairings can be unfavourable due to the phase.
Therefore, both the magnitude and the phase need to be considered or, alternatively, norms such as the RGA-number can be used.
%1971
The Niederlinski index is another measure that only uses steady-state gains of the process model \citep[572-573]{luyben}.
Positive values of the Niederlinski index can correspond to unstable pairings and are therefore inconclusive.
For this reason, the index is considered necessary but not sufficient for stability \citep[445]{skogestad}.
Both the RGA and the Niederlinski index have the advantage of requiring little process information \citep[692]{marlin}.
%1985
\citet{artrdg} defined the RDG, which expanded on the RGA to include the effect of disturbances.
This expansion allowed for the calculation of MV saturation limits and system responses for specified load disturbances \citep{vinsonphd}. 
%1985
Singular value decomposition stems from the use of eigenvalues as an operability measure. 
The use of singular values are preferred as eigenvalues are a poor
measure of gain and can often be misleading \citep[75]{skogestad}.
Maximum and minimum singular values are mostly used to determine good input and output pairings as they characterise the directionality of these pairs.
Along with the condition number, singular values can also be used as a performance measure \citepp{596}{luyben}{80-82}{skogestad}.

@@

The following spaces are of interest:
\begin{description}
  \item [Available Input Space (AIS);] the set of values that process      inputs can reach. 
The limits on these values are based on both physical limits (e.g.     valve openings) or process design values (e.g. flow-rates or temperatures). 
  \item [Achievable Output Space (AOS);] this is the set of values which the process outputs can obtain, given the AIS. 
The AIS maps to the AOS by means of a process model, $G$. 
Therefore, a point $u$ in the AIS corresponds to a point $y$ in the AOS via $y=G(u)$. 
  \item [Desired Output Space (DOS);] this represents the desired output values of the process. 
These values are typically based on operational and financial   parameters. 
  \item [Expected Disturbance Space (EDS);] all the values of the expected disturbances to the system. 
The effects of the disturbances on the inputs are translated back to the input space by means of a disturbance model.
  \item [Desired Input Space (DIS);] in the same manner that the AIS maps to the AOS, the DIS represents combined reverse mappings of the DOS and EDS.
The DIS is calculated as the inputs, $u$, that satisfy the model, i.e. $y=G(u,d)$ (with $y$ and $d$ from the DOS and EDS respectively).
\end{description}

@@

\subsection{Constraint types}\label{sec:contypes}
When constraints are considered, and specifically changes to them, the types of constraints present are important.
This is due to the direction of change allowed by each constraint type.
The following set of constraint types (based on the sources of constraints) are proposed to disambiguate the specification of constraints.
\begin{description}
  \item [Physical constraints] represent the physical limits of the system.
  Examples are tank levels, valve saturation limits and maximum heat duties.
  Allowed changes to these constraints are usually one-sided or to the inside of a range.
  \item [Quality constraints] are concerned with product quality, these constraints are typically determined by external requirements or standards.
  Strictly speaking, changes to these constraints are allowed in any direction.
  \item [Operational constraints] are the innermost constraint set used by the MPC controller.
  These constraints are typically changed using the operator interfaces of MPCs and are contained within their respective variables' physical and quality constraints.
  \item [Optimisation constraints] are those constraints that are not critical for the control of the plant, but rather the performance and efficiency of the plant as determined by external cost functions.
  Typically these external cost functions will represent a cost benefit, e.g. energy savings or product give-away reduction.
\end{description}

@@

\section{Conclusions}\label{sec:conclusions}
From the proposed method of systematic constraint handling and the illustrative example, the conclusions can be drawn:
\begin{itemize}
  \item Checking constraints via the process model can identify constraints which are not attainable and should be revised.
  \item The constraints describing the AOS and DOS intersection represent the largest operating region for the given input and output constraints.
By adding unmeasured variables to the process model, these linear constraints can be reformatted and directly applied in the high and low limit structure of commercial MPC interfaces.
  \item When constraint changes are made (during operation or design), checking changes to constraints via the process model allows for a measure of clamping (or relaxing) to be determined.
This highlights the effect of changing constraints and can guide design or operation decisions.
\end{itemize}
